---
title: ""
format:
  revealjs:
    width: 1080
    height: 1080
    slide-number: false
    transition: none
    center: true
    theme: simple
    css: styles.css
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(ggplot2)
library(data.table)
library(lme4)

source("R/create_dataset.R")
source("R/plots.R")

dataset <- create_dataset()
dataset_bal  <- create_dataset_balanced()

m_lmer_conf <- lmer(
    Yield ~ Temp + Solvent + (1 | Soil),
    data = dataset
    )

m_lmer_bal  <- lmer(
    Yield ~ Temp + Solvent + (1 | Soil),
    data = dataset_bal
    )
```

## Experimental structure

Each soil is observed at only one temperature  
<br>

```{r}
#| label: confounding

make_plot(
    "confounding",
    dt = dataset,
    title = FALSE
    )
```

---

## What the data actually are

48 measurements ≠ 48 independent observations  
<br>

```{r}
#| label: clustering

make_plot(
    "clustering",
    dt = dataset,
    title = FALSE
    )
```

---

## A limitation in the original design

Soils and temperature are confounded
<br>

```{r}
#| label: confounfing

make_plot(
    "confounding",
    dt = dataset,
    title = FALSE
    )
```

---

## Same data. Different structure

What if each soil is tested at all temperatures?  
<br>

```{r}
#| label: structure

make_plot(
    "structure",
    dt_conf = dataset,
    dt_bal = dataset_bal,
    title = FALSE
    )
```

---

## Same data. Different uncertainty

Same number of observations. Different independent information.
<br>

```{r}
#| label: se

make_plot(
    "se",
    m_conf = m_lmer_conf,
    m_bal = m_lmer_bal,
    title = FALSE
    )
```