---
title: "Trying to Understand Variability in Extraction Yield"
format: html
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: libraries_functions

library(knitr)
source("R/create_dataset.R")
```

## Why I started looking into mixed models

I did not start from mixed models.

I started from a practical concern in analytical chemistry:

> how much of the variability I see is due to the method, and how much comes from the matrix?

More specifically:

> am I overestimating the reliability of my results because I am treating dependent observations as independent?

## The experiment

The goal is to study extraction yield of a compound.

We vary:

- extraction solvent  
- temperature  

And we measure:

- extraction yield  

At first glance, this looks like a standard factorial experiment.

## What makes this experiment different

The samples are not identical.

They come from **different soil samples**.

Each soil:

- has its own composition  
- has its own organic content  
- behaves differently during extraction  

Within each soil sample:

- we apply different solvents  
- we vary temperature  
- we take replicate measurements  

So the structure is:

- Soil sample  
  → Solvent / Temperature  
    → Replicates  

## The data

```{r}
#| label: tbl-dataset
#| tbl-cap: "Six random soil samples were extracted at three different temperatures, with three different solvents twice. The yield of a chemical compound was measured in all the samples."

dataset <- create_dataset()
knitr::kable(dataset)
```

## First observation

```{r}
#| label: first_plot

library(ggplot2)

ggplot(dataset, aes(x = Soil, y = Yield)) +
    geom_jitter(width = 0.1, alpha = 0.7) +
    labs(title = "Extraction yield by soil sample") +
    theme_minimal()
```

At this point, something becomes difficult to ignore.

Measurements from the same soil were consistently closer to each other than to measurements from other soils.

That was the first sign that I was not observing independent data points, but repeated observations of the same underlying systems.

## A question that changed the analysis

> are these observations independent?

Because if they are not:

- the model I use is misaligned
- the amount of information is overestimated

## Looking at the variability more carefully

```{r}
#| label: soil_variability

soil_means <- dataset[, .(mean_yield = mean(Yield)), by = Soil]

ggplot(dataset, aes(x = Soil, y = Yield)) +
    geom_point(alpha = 0.4) +
    geom_point(data = soil_means, aes(y = mean_yield), size = 3) +
    labs(title = "Within vs between soil variability") +
    theme_minimal()
```

There are clearly two sources of variation:

- differences between soils
- differences within soils

## What I was doing initially

```{r}
#| label: ols

m_naive <- lm(Yield ~ Temp + Solvent, data = dataset)
summary(m_naive)
```

This assumes:

- all observations are independent
- variability is homogeneous

At this point, that assumption no longer seems reasonable.

## First correction: acknowledging structure

```{r}
#| label: anova

m_aov <- aov(Yield ~ Temp + Solvent + Error(Soil), data = dataset)
summary(m_aov)
```

This respects the experimental design.

But it mainly adjusts how tests are computed.
It still does not parameterize variability as a quantity we can interpret.

## Introducing a mixed model

Instead of asking: what is the correct test?
I started asking: what generates these data?

To answer this question I tried to set up my fist mixel model.

```{r}
#| label: mixed_model

library(lme4)
library(lmerTest)

m_lmer <- lmer(Yield ~ Temp + Solvent + (1 | Soil), data = dataset)
summary(m_lmer)
```

This is where the interpretation changes.

`(1 | Soil)` means:

- each soil has its own baseline
- these baselines vary
- that variability is estimated

## Looking at the variability directly

```{r}
#| label: variability

vc <- VarCorr(m_lmer)

var_soil <- as.numeric(vc$Soil)
var_res  <- attr(vc, "sc")^2

var_soil
var_res
```

Now variability is explicit:

- between soils
- within soils

The estimation of similarity among samples within a soil can be calculated with intraclass correlation (ICC):

```{r}
#| label: ICC

ICC <- var_soil / (var_soil + var_res)
ICC
```

This can be interpreted as the proportion of total variance attributable to differences between soils.
An ICC close to 1 means that observations within the same soil are very similar.

In that case, adding more replicates within the same soil increases precision within that soil,
but contributes very little to understanding variability across soils.

I was optimizing the wrong dimension of my experiment: replication instead of diversity.

## Visualizing what remains within soils

```{r}
#| label: within_soils

dt_centered <- copy(dataset)
dt_centered[, centered := Yield - mean(Yield), by = Soil]

ggplot(dt_centered, aes(x = Soil, y = centered)) +
    geom_point(alpha = 0.6) +
    labs(title = "Within-soil variability after removing soil effects",
    y = "Mean centered yield") +
    theme_minimal()
```

Most of the variation is removed when accounting for soil.

## What I am actually estimating

I am not trying to compare these specific soils.
I am trying to understand how much extraction yield would vary if I analyzed new soils.

This is a property of a population and it is estimated by a so called **random effect**.

To estimate a property of a population, I'm assuming:

- soils are representative
- there are enough of them

Otherwise the estimate of variability may not be reliable

## Exploring uncertainty of the estimation

To explore this, I used a parametric bootstrap based on the fitted model.
This generates new datasets consistent with the model assumptions, including the estimated variability between soils.

The resulting distribution shows that even variance estimates can be quite unstable with few groups.

```{r}
#| label: bootstrap
#| cache: true

set.seed(123)

boot_fun <- function(model) {
  as.numeric(VarCorr(model)$Soil)
}

boot_res <- bootMer(
  m_lmer,
  FUN = boot_fun,
  nsim = 200,
  use.u = TRUE,
  type = "parametric"
)
```

```{r}
#| label: boostrap_plot

ggplot(data.frame(var = boot_res$t), aes(x = var)) +
    geom_histogram(bins = 20) +
    labs(title = "Estimated variability between soils is itself uncertain") +
    theme_minimal()
```

## Implications for experimental design

If the goal is to understand variability across soils:

- adding more replicates within the same soil has diminishing returns
- adding more soils increases the amount of independent information

This was not obvious to me at the beginning.

## What would happen if I ignored soil?

Up to this point, I have two possible models:

- a model that ignores soil structure  
- a model that explicitly accounts for it  

The question is:

> does it actually make a difference?

```{r}
#| label: summary_comparison

summary(m_naive)
summary(m_lmer)
```

The two models are fitted on the same data, but they are based on different assumptions:

- `lm()` assumes all observations are independent
- `lmer()` assumes observations within the same soil are correlated

This affects how uncertainty is estimated.

Looking at standard errors:

```{r}
#| label: se_comparison

coef(summary(m_naive))
coef(summary(m_lmer))
```

In the mixed model, standard errors do not change uniformly: some increase, others decrease. This depends on where each effect is estimated in the experimental structure.

In my case:

- temperature is applied at the soil level  
- solvent is applied within each soil  

This leads to two very different situations.

For temperature:

- the effective number of independent observations is the number of soils  
- uncertainty increases  
- standard errors become larger  

For solvent:

- comparisons are made within the same soil  
- many more observations contribute  
- standard errors become smaller  

So the mixed model is not simply more conservative, it reallocates independent information correctly across effects.

If I ignore soil:

- I treat all 48 observations as independent  
- I underestimate uncertainty for effects applied at the soil level  

If I account for soil:

- I recognize that observations are clustered  
- I use the appropriate level of variability for each effect  

Ignoring structure is not neutral and it leads to:

- overconfidence for some effects  
- distorted comparisons between effects  
- potentially misleading conclusions  

## Final takeaway

Mixed models were not a technical upgrade and hey became necessary because:

- the data were structured
- that structure mattered
- ignoring it changed the conclusions
- I was interested in a property of a population, not in comparing well defined categories.

## What I am still trying to understand

- How many soil samples are enough?
- When does this variability start to affect decisions?
- When do these models change conclusions in practice?
- When is the added complexity actually justified?
